---
title: "CEDEN Query"
author: "Skyler Elmstrom"
date: "1/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R package is currently unavailable without an authenticated user account.

The California Water Board hosts a GitHub repository covering the use of an R package and API to acces the CEDEN database:
https://github.com/CAWaterBoardDataCenter/cedenTools

The full documentation can be found here:
https://github.com/CAWaterBoardDataCenter/cedenTools/blob/master/CEDEN%20Web%20Services%20-%20External%20Web%20Services%20Users%20Guide%20-%200.4.docx

```{r}
install.packages('devtools')
devtools::install_github('daltare/cedenTools')
library(cedenTools)
```

I used the `cedenTools` package to retrieve water quality data from October 2009 - December 2019. We need data with the following parameters:

*  regionalBoard: 5S (Central Valley), 2 (San Francisco Bay)
*  

```{r}
data.download <- ceden_query(service = 'cedenwaterqualityMonitoringStationsList',
                             query_parameters = '"filter":[{"regionalBoard":"5S","sampleDateMin":"10/1/2009","sampleDateMax":"12/31/2019"}]',
                             base_URI = 'cedenwebservices.waterboards.ca.gov')

CEDENBenthicMonitoringStationsList
```

